---
title: "Pollen repartition"
author: "DM"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pollen repartition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(ggplot2)
library(dplyr)
```

# Modeling pollen export

Assuming a population with $M$ males and $F$ females, each male $1, .., m, .. M$ has $\delta_m$ gametes to export. The number of gamete per female is not involved here since the aim of the pollen export is to identify on which females males gametes land.

### Panmictic case

In the panmictic case, for a male ($m$), gametes are on average evenly exported on all females (i.e. follow a multinomial distribution : $N_1, ..., N_F \sim M(\boldsymbol{p}, \delta_m)$, with $\boldsymbol{p} = (p_1, ..., p_K)$ and $p_i = \frac{1}{F} = p, \forall i$).

For each female, the expected number of gametes from male $m$ landing on stigma is given by $E(X_i) = \delta_m p_i$, the variance is given by $var(X_i) = \delta_m p_i (1-p_i)$ and $cov(X_i, X_j) = - \delta_m p_i p_j$ and therefore, in the panmictic case $E(X_i) = \delta_m p, \forall i$, $cov(X_i)=\delta_m p(1-p), \forall i$ and $cov(X_i, X_j) = - \delta_m p^2$.

As gametes amount are usually high (i.e. $\delta_m \gg 0$), pollen repartition is quite deterministic, coefficient of variation is generally below 10% for $\delta_m \gt 1000$ (see Fig. 1A et B). It's important to note here that this variability can't be controlled using a Multinomial distribution, and is almost fully controlled by the number of male gametes ($\delta_m$).

```{r, fig.width=10, fig.height=3, fig.align='center', echo=FALSE}

n = c(1, 5, 10, 25, 50, 100, 250, 500, 1000, 1500, 2500, 5000, 10000)
p = seq(0.05, 0.95, length.out = 20)

cv_table = expand_grid(n, p) %>%
  mutate(esperance = n*p,
         variance = n*p*(1-p),
         cv = sqrt(variance)/esperance)

g1 = ggplot(cv_table, aes(x = n, y = p, fill = cv))+
  geom_tile()+
  scale_x_log10()+
  scale_fill_gradient2(low = "blue", mid="lightblue", high="red", midpoint = 1,
                        labels = scales::percent)+
  labs(fill = "Var. coeff.")+
  theme_bw()

tbl_ = cv_table %>% 
    group_by(n) %>% 
    summarise(max_cv = max(cv, na.rm = T),
              min_cv = min(cv, na.rm = T)) %>% pivot_longer(-n)

g2 = ggplot(tbl_, aes(x = n, y = value, color = name))+
  geom_line()+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10(labels = scales::percent)+
  expand_limits(y = 0)+
  scale_color_discrete(label=c('max(CV)','min(CV)'))+
  labs(y = "Variation coeff.", color = "")

cowplot::plot_grid(g1, g2)

```

However, using Dirichlet distribution in addition to Multinomial allows a finer control on heterogeneity. This Dirichlet-multinomial distribution (a compound distribution) is well defined, and presents interesting properties.

In particular, it's now possible to control for the level of randomness (i.e. $var(X_i)$), meaning that in the panmictic case, one can allows either a fully deterministic pollen repartition (low $\alpha_0$) or a fully random pollen repartition (high $\alpha_0$).

Indeed, whereas the expected number of gametes landing on one stigma for a given male is still given by $E(X_i)=\delta_m p_i$, the variance is now given by : $$var(X_i) = \delta_m p_i (1-p_i)(\frac{\delta_m+\sum a_k}{1+\sum a_k})$$ with $$ p_i = \frac{\alpha_i}{\sum \alpha_k}$$ and min/max cv are now independant of $\delta_m$. Note also than when $\alpha_0 \gg \delta_m$, $(\frac{n+\sum a_k}{1+\sum a_k}) \rightarrow 1$ and variances are similar to the multinomial case (when $\delta_m \rightarrow 1$). 

Thus, overall variance can be control through the $\sum \alpha_k$, usually noted $\alpha_0$. Fig. FFF presents min and max variation coefficients for alternative $\alpha_0$ values. 

```{r, echo = FALSE, fig.width=6, fig.height=3, fig.align='center', warning=FALSE, message=FALSE}

alpha_zero = 10^seq(-6, 4, 1)
n = c(100)
cv_table = expand_grid(n, p, alpha_zero) %>%
  mutate(esperance = n*p,
         variance_dir = n*p*(1-p)*((n*alpha_zero)/(1+alpha_zero)),
         cv = sqrt(variance_dir)/esperance)

tbl_ = cv_table %>% 
    group_by(alpha_zero, n) %>% 
    summarise(max_cv = max(cv, na.rm = T),
              min_cv = min(cv, na.rm = T)) %>% pivot_longer(c(-n, -alpha_zero))

 ggplot(tbl_ %>% filter(n==100), aes(x = alpha_zero, y = value, color = name))+
  geom_line()+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10(labels = scales::percent)+
  scale_color_discrete(label=c('max(CV)','min(CV)'))+
  expand_limits(y = 0)+
  labs(y = "Variation coeff.", color = "")
```

